Viscous Transport in Eroding Porous Media

### 2020-01-02 docs/transport/Response_to_Reviewer_1.tex

\begin{document}

\noindent
Dear Silas Alben,
\\ \\
\noindent
Thank you for handling the manuscript, and we would like to thank the
reviewer for their positive and constructive comments.  The attached
manuscript addresses all the reviewer's comments and the result is what
we believe is a stronger manuscript.  We also note that formatting
issues were resolved in the bibliography. An itemized list of the
changes addressing the reviewer's comments are below.
\\ \\
\noindent
Sincerely,
\\
\noindent
Shang-Huan Chiu, Nick Moore, and Bryan Quaife

\section*{Response to Reviewer 1}

\noindent
\comment{The Barycentric formulae (5.9 \& 5.10) for the first
derivatives of Cauchy integral are incorrect. They are NOT true
barycentric form and straightforward numerical experiments will show
that all accuracies are lost up to the boundary using these formulae,
even if $v(x)$ is computed to high accuracy using the formulae (5.6 \&
5.7). (See e.g.~Sec 3.3 of Barnett et.~al.~2015 for more details.)}
\begin{itemize}
  \item Leading up to equation (5.9) and (5.10), we removed all
    instances where we had previously claimed these quadrature rules
    were in Barycentric form.

  \item After equation (5.10), we now state:
    \begin{quotation}
      \noindent
      Note that $v(x)$ is required to compute $v'(x)$ for both the
      interior and exterior case, and this is available using the
      Barycentric quadrature rules (5.6) and (5.7). As pointed out by
      Barnett {\em et al.}~(2015), equations (5.9) and (5.10) are not in
      a Barycentric form since the numerators suffer from catastrophic
      cancellation as $x$ approaches $\partial\Omega$. The result is
      that the quadrature error is not uniformly bounded. However, since
      these errors are sufficiently small for the geometries and
      resolutions we consider, we use these quadrature formulas
      nonetheless. We note that the quadrature formulas can be put into
      true Barycentric form (Barnett {\em et al.}~2015, see equations
      (3.11) and (3.13)), but this increases the complexity for
      evaluating $v'(x)$ to $O(N^2)$ operations. 
    \end{quotation}

\noindent
\comment{Likewise, the second derivative ``barycentric formulae" (5.15 \& 5.18) are incorrect, either. So the authors will need to explain how they managed to get high accuracies with the stated formulae?}
\begin{itemize}
  \item Leading up to equation (5.15) and (5.18), we removed all
    instances where we had previously claimed these quadrature rules
    were in Barycentric form.

  \item We have also updated the text so that it does not give the
    impression that the error is uniformly bounded. However, for the
    geometries and resolutions we consider, we found that the
    non-Barycentric form quadrature rules are accurate enough to compute
    the necessary quantities at the target points that are required to
    perform the subsequent analysis in the paper.

  \item After equation (5.18), we now state:
    \begin{quotation}
      \noindent
      These quadrature formula are also not in a Barycentric form
      because of catastrophic cancellation in the numerator. As is the
      case when computing $v′(x)$, this error is sufficiently small at
      the geometries and resolutions we consider, so we use these
      quadrature formulas.
    \end{quotation}

  \item To make it clear which quadrature methods we apply, we have
    added the following paragraph to the end of Section 5.2:
    \begin{quotation}
      \noindent
      In section 6, the quadrature rules (5.6), (5.7), (5.9), (5.10),
      (5.15), and (5.18) are used to form simulations of nearly-touching
      eroding grains, and to study dynamics of the flow in regions
      arbitrarily close to eroding grains.
    \end{quotation}

  \item In the second paragraph of Section 5.2, we updated the last
    sentence to read:
    \begin{quotation}
      \noindent
      After briefly summarizing this method, we compute the second
      derivative so that the shear stress and vorticity can be
      accurately computed.
    \end{quotation}

  \item We start Section 5.3 with the following revised sentence:
    \begin{quotation}
      \noindent
      By using these quadrature rules, the velocity, shear stress, and
      vorticity are calculated at the target points necessary to
      simulate erosion and transport.
    \end{quotation}

\noindent
\comment{In my opinion, the claim of having developed a ``new quadrature
method" is too strong. After all, there is nothing new about the
barycentric quadrature itself. Rather, I suggest that maybe a more
appropriate claim would be that the authors have developed a ``new
numerical scheme" for the Stokes porous media flow, where the accurate
quadrature is one ingredient in this scheme.}
\begin{itemize}
  \item We have toned down some of the claims. The following modified
    paragraphs and sentences are in the revised manuscript:
    \begin{quotation}
      \noindent
      In previous work (Quaife \& Moore 2018), the BIE is discretized
      with the trapezoid rule, and this is acceptable if the grains that
      are sufficiently separated (Trefethen \& Weideman 2014). However,
      when grains are close, the result is a nearly-singular integrand
      and the error of the trapezoid rule is no longer acceptable. One
      of the earliest quadrature methods for nearly-singular integrands
      was developed by Baker \& Shelley (1986), and in recent years,
      many other schemes have followed (af Klinteberg \& Tornberg 2018;
      Helsing \& Ojala 2008; Beale {\em et al.}~2016; Beale \& Lai 2001;
      Kl\"{o}ckner {\em et al.}~2013). We use a Barycentric quadrature
      method (Barnett 2014; Barnett {\em et al.}~2015) since it is a
      non-intrusive modification of the trapezoid rule, and the error is
      guaranteed to be uniformly bounded. We use this quadrature method
      to compute the velocity gradient, which is needed to evaluate the
      shear stress and the fluid vorticity.
    \end{quotation}

    \begin{quotation}
      \noindent
      Section 5 describes the numerical methods, with special attention
      paid to the quadrature method for computing the shear stress and
      the vorticity.
    \end{quotation}

    \begin{quotation}
      \noindent
      As a continuation of our previous work (Quaife \& Moore 2018), we
      have simulated dense suspensions and characterized transport in
      viscous eroding porous media. This is accomplished by using
      high-order time stepping methods and a combination of trapezoid
      and Barycentric quadrature methods to solve a BIE formulation of
      the Stokes equations. By using these numerical methods, we are
      able to perform stable simulations of erosion with $N = O(100)$
      discretization points, while the trapezoid rule would require
      $O(10^5)$ discretization points.
    \end{quotation}
\end{itemize}

@@

    \begin{quotation}
      \noindent
      Section 5 describes the numerical methods, with special attention
      paid to the quadrature method for computing the shear stress and
      the vorticity.
    \end{quotation}

    \begin{quotation}
      \noindent
      As a continuation of our previous work (Quaife \& Moore 2018), we
      have simulated dense suspensions and characterized transport in
      viscous eroding porous media. This is accomplished by using
      high-order time stepping methods and a combination of trapezoid
      and Barycentric quadrature methods to solve a BIE formulation of
      the Stokes equations. By using these numerical methods, we are
      able to perform stable simulations of erosion with $N = O(100)$
      discretization points, while the trapezoid rule would require
      $O(10^5)$ discretization points.
    \end{quotation}
\end{itemize}

\noindent
\comment{The transient increase of tortuosity is interesting, but the
explanation in Sec 6.4 is not satisfying. The authors proposed that the
increase of tortuosity can be explained by the sudden increase of
$x$-velocity when a grain vanishes. But in formula (4.4) or (4.5), the
$x$-velocity appears both in the numerator and the denominator, which
seems to me that the increase would just cancel. Then, as the authors
have pointed out, wouldn't the straighter streamlines resulted from the
vanishing of grains have decreased the tortuosity?}
\begin{itemize}
  \item response
\end{itemize}

\noindent
\comment{The porosity-tortuosity models (4.6c \& 4.6d) are not
explicitly compared in the results section. How much worse are they
comparing to the other models?}
\begin{itemize}
  \item We have adjusted each of the appropriate sections.

  \item In Section 6.2, we have added the following sentence:
    \begin{quotation}
      \noindent
      A slightly larger root-mean-square error of $6.70 \times 10^{-3}$
      is possible with the logarithmic porosity-tortuosity model (4.6b),
      but the other two models have a root- mean-square error that is an
      order of magnitude larger.
    \end{quotation}

  \item In Section 6.3, we have added the following sentence:
    \begin{quotation}
      \noindent
      However, the root-mean-square error of the other three models is
      bounded by $2.38 \times 10^{-2}$.
    \end{quotation}

  \item In Section 6.4, we have added the following sentence:
    \begin{quotation}
      \noindent
      We also compute the lines of best fit using the
      porosity-tortuosity models (4.6).  Each of the models result in a
      root-mean-squared error ranging from $4.60 \times 10^{-3}$ for
      model (4.6c) to $7.20 \times 10^{-3}$ for model (4.6c). The black
      dashed line in figure 15(a) is the line of best fit
      $\widehat{T}(\phi) = \phi^{-0.2459}$, with a root-mean-square
      error of $5.50 \times 10^{-3}$.  We note a slightly better
      root-mean-square error of $4.60 \times 10^{-3}$ is possible with
      the model $\widehat{T}(\phi) = \left(1 + 0.1499(1-\phi)\right)^2$.
    \end{quotation}

@@

\noindent
\comment{Should the shear stress formula (2.2) contain viscosity in it
based on the governing equation (2.1)?}
\begin{itemize}
  \item In general, it would. However, since we have non-dimensionalized
    following our previous work, the viscosity is assumed to be 1.
\end{itemize}

\noindent
\comment{In Figure 5, do the colors represent vorticity itself or the
log of the abs vorticity? (Which would be consistent with the colors in
Figure 4.)}
\begin{itemize}
  \item The color represents the logarithm of the absolute value of the
    vorticity.
  \item We have updated the caption to read:
    \begin{quotation}
    \noindent
      The eroded bodies from figure 4 at time $t = 0.1$. The color is
      the logarithm of the magnitude of the vorticity. The initial
      distances from the body to the solid wall are: (a) $h$, (b) $h/2$,
      and (c) $h/10$.
    \end{quotation}
  \item We have also updated the text to read:
    \begin{quotation}
    \noindent
      To visualize the flow, in figure 5 we plot the logarithm of the
      absolute value of the vorticity from the final time step in figure
      4.
    \end{quotation}
\end{itemize}

\noindent
\comment{When first introduced in Sec 4.3, the ``pore sizes" are not
clearly defined. (My first impression was that pore size = portion of an
edge of the Delaunay diagram. Only at the end in Sec 6.4 do we know that
it is defined as the closest distance between grains.)}
\begin{itemize}
  \item In Figure 3, we clarify the definition by adding the statement:
    {\em represented by the length of the red line segments}.
  \item In the text, we now state:
    \begin{quotation}
    \noindent
      Finally, we define the pore sizes to be the minimum distance
      between all pairs of neighboring grains, and these distances are
      represented by the red line segments in figure 3(a).
    \end{quotation}
\end{itemize}

\noindent
\comment{``Porosity" is undefined throughout the paper.}
\begin{itemize}
  \item We have now formally defined the porosity by rewriting the
    description of coarse-grained models as:
  \begin{quotation}
  \noindent
    Examples of coarse-grained models for porous media flow include
    tortuosities (Matyka {\em et al.}~2008; Duda {\em et al.}~2011;
    Koponen {\em et al.}~1996), geometry connectivity (Knudby \& Carrera
    2005), anomalous dispersion (Dentz {\em et al.}~2004), and
    permeability-porosity relationships (Dardis \& McCloskey 1998;
    Carman 1937), where the porosity is the fraction of the geometry
    that is fluidized.
  \end{quotation}

### 2020-01-02 docs/transport/erosion_transport.tex
working on revisions

To compute stable simulations of erosion, we use methods of high-order
in both space and time. The time integration is unchanged from previous
work~\citep{qua-moo2018}.  We apply a mild regularization and a
smoothing term to eliminate numerical instabilities that can be
triggered by changes in sign of the shear stress, and we use a stable
second-order Runge-Kutta method applied to the {\thL}
coordinates~\citep{hou-low-she1994} of the eroding grains.  In this
work, we introduce a new quadrature method to resolve dense suspensions.
The accuracy of the trapezoid rule, which was used in previous work, is
adequate for bodies that are sufficiently separated~\citep{tre-wei2014},
but not for grains in close contact.  One of the earliest quadrature
methods for nearly-singular integrands was developed
by~\citet{bak-she1986}, and in recent years, many other schemes have
followed~\citep{kli-tor2018, hel-oja2008a, bea-yin-wil2016, bea-lai2001,
klo-bar-gre-one2013}.  We use a Barycentric quadrature
method~\citep{bar2014, bar-wu-vee2015} since it is a non-intrusive
modification of the trapezoid rule, and the error is guaranteed to be
uniformly bounded.  We extend the original quadrature method to compute
the velocity gradient, which is needed to evaluate the shear stress and
the fluid vorticity.

While natural porous media are three-dimensional, there are several
challenges that must be overcome to extend our high-order methods to
three dimensions.  A three-dimensional version of a {\thL} formulation
exists~\citep{amb-sie-tlu2013}, but the Barycentric quadrature method
relies upon a complex variable formulation of the BIE, making it
inherently two-dimensional. Therefore, a different quadrature method is
required. Additional challenges of three-dimensional simulations include
spectral representations of the geometry and efficiently solving
discretizations of the BIE. While high-order methods for
three-dimensional BIEs have been developed, their computational cost
limits the number of bodies that can be practically simulated. Another
limitation of this work is the assumption of immobile grains.
Mechanisms such as sedimentation play a role in real applications, and
this will effect the transport properties. Simulating mobile grains
introduces the additional challenge of resolving contact.  Since
high-order methods have never been used to study transport in eroded
porous media, as a first step we consider the case  with two-dimensional
immobile grains. This assumption allows for a clear interpretation of
the results.  For example, in this work we are able to simulate a large
densely-packed suspension (up to $100$ in this manuscript).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
This paper is organized as follows. In section~\ref{sec:formulation}, we
summarize the erosion model that is described in more detail in previous
work~\citep{qua-moo2018}.  In section~\ref{sec:DLP}, we recast all the
governing equations as layer potentials defined in both $\RR^2$ and in
$\CC$.  Section~\ref{sec:transport} describes measures for
characterizing the geometry and transport. Section~\ref{sec:method}
describes the numerical methods, with special attention paid to the new
quadrature method for computing the shear stress and the vorticity.
Section~\ref{sec:results} presents numerical examples for a variety of
dense packings of bodies.  Finally, concluding remarks are made in
section~\ref{sec:conclusions}.

@@

There have been efforts to relate the tortuosity, $T$, to the porosity,
$\phi$.  For example,~\citet{mat-kha-koz2008} propose the
models
\begin{subequations}
  \label{eqn:tortuosityModels}
  \begin{align}
    \widehat{T}(\phi) &= \phi^{-p}, \\
    \widehat{T}(\phi) &= 1-p \log \phi, \\
    \widehat{T}(\phi) &= 1+p (1-\phi), \\
    \widehat{T}(\phi) &= (1+p (1-\phi))^2, 
  \end{align}
\end{subequations}
where $p>0$ is a fitting parameter.  In section~\ref{sec:results}, we
compare these four models with the tortuosity of eroding porous
geometries.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Barycentric Quadrature Formulas}
\label{sec:bary}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
While the trapezoid rule is spectrally accurate for smooth, periodic
functions~\citep{tre-wei2014}, the derivative of the integrand grows as
the target point $\xx$ approaches $\bd\Omega$.  Therefore, if the
trapezoid rule is applied when bodies are in near-contact, or if a layer
potential is evaluated at a point close to $\bd\Omega$, then the result
become unreliable and the simulation ultimately becomes unstable.  We
thus desire a quadrature method whose error bound does not depend on the
target location.

We showed in sections~\ref{sec:DLPcomplex} and~\ref{sec:gradDLPcomplex}
that the velocity, shear stress, and vorticity of the double-layer
representation can all be written as the sum of Cauchy integrals and its
first two derivatives.  Therefore, we require quadrature rules with a
uniform error bound for Cauchy integrals and its
derivatives~\eqref{eqn:cauchy}. \citet{ioa-pap-per1991} developed
quadrature rules, that we call {\em Barycentric quadrature rules}, to
compute Cauchy integrals and their derivatives with an error bound that
is independent of $x \in \CC$.  Then,~\citet{bar-wu-vee2015} used these
quadrature rules to compute the Stokes double-layer potential
representation of the velocity~\eqref{eqn:velocityDLP}. After briefly
summarizing this method, we extend the work to compute the second
derivative so that the shear stress and vorticity can be computed with a
uniform error bound.

@@

Similar constructions can be used to derive quadrature rules
for $v'(x)$.  For $x \in \Omega$, we have the identity
\begin{align}
  \int_{\bd\Omega} \frac{v^-(y) - v(x) + (y-x)v'(x)}{(y - x)^2} = 0,
\end{align}
and the integrand is bounded for all $x \in \Omega$.  Therefore, after
applying the trapezoid rule and rearranging for $v'(x)$, we have the
interior Barycentric quadrature rule
\begin{align}
  v'(x) = \left(\sum_{j=1}^{N}
    \frac{v^{-}(y_j) - v(x)}{(y_j-x)^2} w_j \right)
  \Bigg/
  \left(\sum_{j=1}^{N} \frac{1}{y_j-x} w_j\right), 
  \quad x \in \Omega.
  \label{eqn:BaryvprimeInterior}
\end{align}
Using a similar construction, the exterior Barycentric quadrature rule
for the first derivative is
\begin{align}
  v'(x) = \frac{1}{x-a} \left(\sum_{j=1}^N
    \frac{v^+_j - v(x)}{(y_j - x)^2} w_j \right)
    \Bigg/
    \left(\sum_{j=1}^N \frac{(y_j-a)^{-1}}{y_j - x} w_j\right),
    \quad x \in \Omega^c.
  \label{eqn:BaryvprimeExterior}
\end{align}
Note that $v(x)$ is required to compute $v'(x)$ for both the interior
and exterior case, and this is available using the Barycentric
quadrature rules~\eqref{eqn:BaryvInterior}
and~\eqref{eqn:BaryvExterior}.

To compute the shear stress and vorticity, we require a Barycentric
quadrature rule for $v''(x)$.  The derivation is largely based on the
work of~\citet[see equation (2.12)]{ioa-pap-per1991}.  We start with
the second derivative of the Cauchy integral theorem
\begin{align}
  0 = \frac{1}{2\upi i} \int_{\bd\Omega} 
      \frac{2v^{-}(y)}{(y-x)^3}\,dy - v''(x).
\end{align}
For the interior case, $x \in \Omega$, we use the identity
\begin{align}
  \frac{1}{2\upi i}\int_{\bd\Omega} \frac{1}{(y-x)^n}\, dy = 
  \left\{
    \begin{array}{ll}
      1, & n = 1, \\
      0, & n = 2,3,\ldots. \\
    \end{array}
  \right.
\end{align}
Combining this identity with the Cauchy integral representation of
$v''(x)$, we have
\begin{align}
  0 = \frac{1}{2\upi i} \int_{\bd\Omega} 
      \frac{2v^{-}(y) - v''(x)(y-x)^2 - 2v(x) - 2(y-x)v'(x)}
      {(y-x)^3}\, dy.
\end{align}
This integrand is constructed so that it is bounded for all $x \in
\Omega$, and applying the trapezoid rule, we have
\begin{align}
  0 \approx  \sum_{j=1}^{N} 
      \frac{2v^{-}(y_j) - v''(x)(y_j-x)^2 - 2v(x) - 2(y_j-x)v'(x)}
      {(y_j-x)^3} w_j,
\end{align}
where the accuracy is independent of $x$.  Solving for $v''(x)$, the
Barycentric quadrature rule for the interior second derivative at $x \in
\Omega$ is
\begin{align}
  v''(x) \approx \left(2\sum_{j=1}^N 
    \frac{v^{-}_{j} - v(x) - (y_j-x)v'(x)}{(y_j-x)^3}w_j \right)
    \Bigg/
    \left(\sum_{j=1}^N \frac{1}{y_j-x}w_j\right).
\end{align}
For the exterior case, $x \in \Omega^c$, we start with the identity
\begin{align}
\frac{1}{x-a} &= -\frac{1}{2\upi i}\int_{\bd\Omega} 
    \frac{(y-a)^{-1}}{y-x}\, dy. 
\end{align}
Combining this identity with the Cauchy integral representation of
$v''(x)$, we have
\begin{align}
  0 = \frac{1}{2\upi i} \int_{\bd\Omega} 
    \frac{2v^+(y) - 2v(x) - 2(y-x)v'(x) + (y-a)^{-1} (x-a) (y-x)^2 v''(x)}
    {(y-x)^3}.
\end{align}
As in the interior case, the integrand is chosen so that it is bounded
for all $x \in \Omega^c$.  Therefore, after applying the trapezoid rule
and solving for $v''(x)$, we have the Barycentric quadrature rule for
the exterior second derivative at $x \in \Omega^c$
\begin{align}
  v''(x) \approx \frac{1}{x-a}\left(2\sum\limits_{j=1}^N
    \frac{v^{+}_{j} - v(x) - (y_j-x)v'(x)}{(y_j-x)^3}w_j\right)
    \Bigg/
    \left(\sum_{j=1}^N \frac{(y_j-a)^{-1}}{y_j-x}w_j\right).
\end{align}
The quadrature rule for $v''(x)$ requires $v(x)$ and $v'(x)$, and theses
are computed using the quadrature rules in
equations~\eqref{eqn:BaryvInterior},~\eqref{eqn:BaryvExterior},~\eqref{eqn:BaryvprimeInterior},
and~\eqref{eqn:BaryvprimeExterior}. In section~\ref{sec:results}, these
quadrature rules are used to form simulations of nearly-touching eroding
grains, and to study dynamics of the flow in regions arbitrarily close
to eroding grains.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Efficiently Applying the Quadrature}
\label{sec:fmm}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
By using the Barycentric quadrature rule, the velocity, shear stress,
and vorticity are computed with an error that is bounded independent of
the target location. However, applied directly, it requires
$O(N^2)$ operations, where $N$ is the total number of source
and target points.  By using a fast summation method, such as the fast
multipole method (FMM)~\citep{gre-rok1987}, the cost can be reduced to
$O(N)$ operations.  However, each application of the Barycentric
quadrature rules involve several $N$-body calculations, rendering the
computational cost prohibitive, so we introduce a hybrid method that
combines the Barycentric quadrature rule and an accelerated trapezoid
rule.  Note that the source points of the layer potential is always one
of the eroding bodies or the outer wall, but the target point can either
be on another component of $\bd\Omega$ or it can be in the fluid bulk
$\Omega$.

>>>>

To compute stable simulations of erosion, we use methods of high-order
in both space and time. The time integration is unchanged from previous
work~\citep{qua-moo2018}.  We apply a mild regularization and a
smoothing term to eliminate numerical instabilities that can be
triggered by changes in sign of the shear stress, and we use a stable
second-order Runge-Kutta method applied to the {\thL}
coordinates~\citep{hou-low-she1994} of the eroding grains. In previous
work~\citep{qua-moo2018}, the BIE is discretized with the trapezoid
rule, and this is acceptable if the grains that are sufficiently
separated~\citep{tre-wei2014}. However, when grains are close, the
result is a nearly-singular integrand and the error of the trapezoid
rule is no longer acceptable. One of the earliest quadrature methods for
nearly-singular integrands was developed by~\citet{bak-she1986}, and in
recent years, many other schemes have followed~\citep{kli-tor2018,
hel-oja2008a, bea-yin-wil2016, bea-lai2001, klo-bar-gre-one2013}.  We
use a Barycentric quadrature method~\citep{bar2014, bar-wu-vee2015}
since it is a non-intrusive modification of the trapezoid rule, and the
error is guaranteed to be uniformly bounded.  We use this quadrature
method to compute the velocity gradient, which is needed to evaluate the
shear stress and the fluid vorticity.

Since high-order methods have never been used to study transport in
eroding porous media, we consider, as a first step, two-dimensional and
immobile grains. These assumptions allow us to simulate large
densely-packed suspensions (up to $100$ in this manuscript), and to
clearly interpret the results. More realistic three-dimensional and
mobile configurations would require overcoming a few challenges. First,
a three-dimensional extension of the {\thL} formulation
exists~\citep{amb-sie-tlu2013}, but the Barycentric quadrature method
used here relies on a complex variable formulation of the BIE, making it
inherently two-dimensional. Therefore, a different quadrature method
would be required. Additional challenges include spectral
representations of three-dimensional geometries and efficiently solving
discretizations of the BIE. While high-order methods for
three-dimensional BIEs have been developed, their computational cost
limits the number of bodies that can be practically simulated.  Lastly,
mechanisms such as sedimentation play a role in realistic media and
affect the pore structure. Simulating mobile grains introduces the
additional challenge of resolving contact, which we plan to address in
future work.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
This paper is organized as follows. In section~\ref{sec:formulation}, we
summarize the erosion model that is described in more detail in previous
work~\citep{qua-moo2018}.  In section~\ref{sec:DLP}, we recast all the
governing equations as layer potentials defined in both $\RR^2$ and in
$\CC$.  Section~\ref{sec:transport} describes measures for
characterizing the geometry and transport. Section~\ref{sec:method}
describes the numerical methods, with special attention paid to the
quadrature method for computing the shear stress and the vorticity.
Section~\ref{sec:results} presents numerical examples for a variety of
dense packings of bodies.  Finally, concluding remarks are made in
section~\ref{sec:conclusions}.

@@

There have been efforts to relate the tortuosity, $T$, to the porosity,
$\phi$.  For example,~\citet{mat-kha-koz2008} propose the models
\begin{subequations}
  \label{eqn:tortuosityModels}
  \begin{align}
    \label{eqn:tortuosityModels1}
    \widehat{T}(\phi) &= \phi^{-p}, \\
    \label{eqn:tortuosityModels2}
    \widehat{T}(\phi) &= 1-p \log \phi, \\
    \label{eqn:tortuosityModels3}
    \widehat{T}(\phi) &= 1+p (1-\phi), \\
    \label{eqn:tortuosityModels4}
    \widehat{T}(\phi) &= (1+p (1-\phi))^2, 
  \end{align}
\end{subequations}
where $p>0$ is a fitting parameter.  In section~\ref{sec:results}, we
compare these four models with the tortuosity of eroding porous
geometries.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Barycentric Quadrature Formulas}
\label{sec:bary}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
While the trapezoid rule is spectrally accurate for smooth, periodic
functions~\citep{tre-wei2014}, the derivative of the integrand grows as
the target point $\xx$ approaches $\bd\Omega$.  Therefore, if the
trapezoid rule is applied when bodies are in near-contact, or if a layer
potential is evaluated at a point close to $\bd\Omega$, then the result
becomes unreliable and the simulation ultimately becomes unstable.  We
thus desire a quadrature method whose error bound does not depend on the
target location.

We showed in sections~\ref{sec:DLPcomplex} and~\ref{sec:gradDLPcomplex}
that the velocity, shear stress, and vorticity of the double-layer
representation can all be written as the sum of Cauchy integrals and its
first two derivatives.  Therefore, we require quadrature rules with a
uniform error bound for Cauchy integrals and its
derivatives~\eqref{eqn:cauchy}. \citet{ioa-pap-per1991} developed
quadrature rules, that we call {\em Barycentric quadrature rules}, to
compute Cauchy integrals and their derivatives with an error bound that
is independent of $x \in \CC$.  Then,~\citet{bar-wu-vee2015} used these
quadrature rules to compute the Stokes double-layer potential
representation of the velocity~\eqref{eqn:velocityDLP}. After briefly
summarizing this method, we compute the second derivative so that the
shear stress and vorticity can be accurately computed.

@@

Similar constructions can be used to derive quadrature rules
for $v'(x)$.  For $x \in \Omega$, we have the identity
\begin{align}
  \int_{\bd\Omega} \frac{v^-(y) - v(x) + (y-x)v'(x)}{(y - x)^2} = 0,
\end{align}
and the integrand is bounded for all $x \in \Omega$.  Therefore, after
applying the trapezoid rule and rearranging for $v'(x)$, we have the
interior quadrature rule
\begin{align}
  v'(x) = \left(\sum_{j=1}^{N}
    \frac{v^{-}(y_j) - v(x)}{(y_j-x)^2} w_j \right)
  \Bigg/
  \left(\sum_{j=1}^{N} \frac{1}{y_j-x} w_j\right), 
  \quad x \in \Omega.
  \label{eqn:BaryvprimeInterior}
\end{align}
Using a similar construction, an exterior quadrature rule for the first
derivative is
\begin{align}
  v'(x) = \frac{1}{x-a} \left(\sum_{j=1}^N
    \frac{v^+_j - v(x)}{(y_j - x)^2} w_j \right)
    \Bigg/
    \left(\sum_{j=1}^N \frac{(y_j-a)^{-1}}{y_j - x} w_j\right),
    \quad x \in \Omega^c.
  \label{eqn:BaryvprimeExterior}
\end{align}
Note that $v(x)$ is required to compute $v'(x)$ for both the interior
and exterior case, and this is available using the Barycentric
quadrature rules~\eqref{eqn:BaryvInterior}
and~\eqref{eqn:BaryvExterior}. As pointed out by~\citet{bar-wu-vee2015},
equations~\eqref{eqn:BaryvprimeInterior}
and~\eqref{eqn:BaryvprimeExterior} are not in a Barycentric form since
the numerators suffer from catastrophic cancellation as $x$ approaches
$\bd\Omega$. The result is that the quadrature error is not uniformly
bounded. However, since these errors are sufficiently small for the
geometries and resolutions we consider, we use these quadrature formulas
nonetheless. We note that the quadrature formulas can be put into true
Barycentric form~\citep[see equations (3.11) and
(3.13)]{bar-wu-vee2015}, but this increases the complexity for
evaluating $v'(x)$ to $O(N^2)$ operations.

To compute the shear stress and vorticity, we require a quadrature rule
for $v''(x)$.  The derivation is largely based on the work of~\citet[see
equation (2.12)]{ioa-pap-per1991}.  We start with the second derivative
of the Cauchy integral theorem
\begin{align}
  0 = \frac{1}{2\upi i} \int_{\bd\Omega} 
      \frac{2v^{-}(y)}{(y-x)^3}\,dy - v''(x).
\end{align}
For the interior case, $x \in \Omega$, we use the identity
\begin{align}
  \frac{1}{2\upi i}\int_{\bd\Omega} \frac{1}{(y-x)^n}\, dy = 
  \left\{
    \begin{array}{ll}
      1, & n = 1, \\
      0, & n = 2,3,\ldots. \\
    \end{array}
  \right.
\end{align}
Combining this identity with the Cauchy integral representation of
$v''(x)$, we have
\begin{align}
  0 = \frac{1}{2\upi i} \int_{\bd\Omega} 
      \frac{2v^{-}(y) - v''(x)(y-x)^2 - 2v(x) - 2(y-x)v'(x)}
      {(y-x)^3}\, dy.
\end{align}
This integrand is constructed so that it is bounded for all $x \in
\Omega$, and applying the trapezoid rule, we have
\begin{align}
  0 \approx  \sum_{j=1}^{N} 
      \frac{2v^{-}(y_j) - v''(x)(y_j-x)^2 - 2v(x) - 2(y_j-x)v'(x)}
      {(y_j-x)^3} w_j,
\end{align}
where the accuracy is independent of $x$.  Solving for $v''(x)$, the
quadrature rule for the interior second derivative at $x \in
\Omega$ is
\begin{align}
  v''(x) \approx \left(2\sum_{j=1}^N 
    \frac{v^{-}_{j} - v(x) - (y_j-x)v'(x)}{(y_j-x)^3}w_j \right)
    \Bigg/
    \left(\sum_{j=1}^N \frac{1}{y_j-x}w_j\right).
    \label{eqn:Baryvprime2Interior}
\end{align}
For the exterior case, $x \in \Omega^c$, we start with the identity
\begin{align}
\frac{1}{x-a} &= -\frac{1}{2\upi i}\int_{\bd\Omega} 
    \frac{(y-a)^{-1}}{y-x}\, dy. 
\end{align}
Combining this identity with the Cauchy integral representation of
$v''(x)$, we have
\begin{align}
  0 = \frac{1}{2\upi i} \int_{\bd\Omega} 
    \frac{2v^+(y) - 2v(x) - 2(y-x)v'(x) + (y-a)^{-1} (x-a) (y-x)^2 v''(x)}
    {(y-x)^3}.
\end{align}
As in the interior case, the integrand is chosen so that it is bounded
for all $x \in \Omega^c$.  Therefore, after applying the trapezoid rule
and solving for $v''(x)$, we have the quadrature rule for the exterior
second derivative at $x \in \Omega^c$
\begin{align}
  v''(x) \approx \frac{1}{x-a}\left(2\sum\limits_{j=1}^N
    \frac{v^{+}_{j} - v(x) - (y_j-x)v'(x)}{(y_j-x)^3}w_j\right)
    \Bigg/
    \left(\sum_{j=1}^N \frac{(y_j-a)^{-1}}{y_j-x}w_j\right).
    \label{eqn:Baryvprime2Exterior}
\end{align}
The quadrature rule for $v''(x)$ requires $v(x)$ and $v'(x)$, and these
are computed using the quadrature rules in
equations~\eqref{eqn:BaryvInterior},~\eqref{eqn:BaryvExterior},~\eqref{eqn:BaryvprimeInterior},
and~\eqref{eqn:BaryvprimeExterior}. These quadrature formula are also
not in a Barycentric form because of catastrophic cancellation in the
numerator. As is the case when computing $v'(x)$, this error is
sufficiently small at the geometries and resolutions we consider, so we
use these quadrature formulas.

In section~\ref{sec:results}, the quadrature
rules~\eqref{eqn:BaryvInterior},~\eqref{eqn:BaryvExterior},~\eqref{eqn:BaryvprimeInterior},~\eqref{eqn:BaryvprimeExterior},~\eqref{eqn:Baryvprime2Interior},
and~\eqref{eqn:Baryvprime2Exterior} are used to form simulations of
nearly-touching eroding grains, and to study dynamics of the flow in
regions arbitrarily close to eroding grains.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Efficiently Applying the Quadrature}
\label{sec:fmm}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
By using these quadrature rules, the velocity, shear stress, and
vorticity are calculated at the target points necessary to simulate
erosion and transport. However, applied directly, it requires $O(N^2)$
operations, where $N$ is the total number of source and target points.
By using a fast summation method, such as the fast multipole method
(FMM)~\citep{gre-rok1987}, the cost can be reduced to $O(N)$ operations.
However, each application of the Barycentric quadrature rules involve
several $N$-body calculations, rendering the computational cost
prohibitive, so we introduce a hybrid method that combines the
Barycentric quadrature rule and an accelerated trapezoid rule.  Note
that the source points of the layer potential is always one of the
eroding bodies or the outer wall, but the target point can either be on
another component of $\bd\Omega$ or it can be in the fluid bulk
$\Omega$.
